Parameterization of the responses of subarctic European vegetation to key environmental variables for ozone risk assessment

### 2022-05-31 VegOzone.tex
Fine tuning and comments by Lisa

\subsubsection{Gap filling methodology}
\label{subsec:gap_filling}
The $\mathrm{DO_3SE}$ model requires hourly, continuous input data. Therefor, the following gap filling methodology was devised for the meteorological drivers of the model:
\begin{itemize}
\item Single hours of missing data were filled by taking the average of the hourly values before, and after, the missing value.
\item Several consecutive hours of missing data (23 or less) were filled by taking the average of the corresponding hour the day before, and the day after; and repeating this for each missing hour of data. If data were unavailable from that hour of the previous day, then only the value from the day after was used and vice versa. Interpolated values were not used in calculating averages.
\item Data gaps longer than $24\,\unit{h}$ were filled using weekly diurnal averages. i.e. an average was calculated using the corresponding hour throughout the week before and after. Interpolated values were not used in calculating averages.
\end{itemize}

@@

\subsubsection{Input data and observation site}
\label{subsec:data}
The required input data consists of ozone concentration $\chem{[O_3]}$, $2\,\unit{m}$ temperature $T_\mathrm{2m}$, global irradiation $Q_0$, precipitation, $2\,\unit{m}$ wind $u_\mathrm{2m}$, and vapor pressure deficit VPD. VPD has been calculated from observed $T_\mathrm{2m}$ and relative humidity
\begin{equation}
  \mathrm{VPD} = P_s(T) \cdot \left(1-\frac{\mathrm{relHum}}{100}\right),
\end{equation}
with saturation vapor pressure $P_s(T)$ in \unit{hPa} from the Arden~Buck~equation~\citep{JAP:Buck1981, Buck2012}.

For our study, we chose Svanhovd in northern Norway ($69.45\,\unit{^\circ N}$, $30.03\,\unit{^\circ E}$, $30\,\unit{m\,a.s.l.}$). Reasons for our choice were: (1) Ambient ozone concentration had been monitored there in the past, (2) an agrometeorological station is measuring, e.g. $2\,\unit{m}$ temperature, precipitation, global irradiance, and soil temperature at different depths, and (3) a national surveillance of air pollution (\chem{NO_x}, \chem{SO_x}, particular matter (PM), radiological species) is ongoing. Most of these measurements are performed by the Norwegian Institute for Air Research (NILU), who also conducted ozone measurements on our behalf during the 2018/19 growing season.

All essential climate variables for the 2018/19 growing season were provided by NILU through the web portal \href{luftkvalitet.no}{luftkvalitet.no} (last accessed May 2020). The ozone monitoring data have been added to a database for air-quality monitoring called EBAS as well operated by NILU and are thus openly accessible. Long-term ozone observation data (1986--1996) were obtained from EBAS. Note that data from before the 1990s do not follow the high-quality standard procedures implemented nowadays and have to be treated with care \citep{NILU2003}. Because ozone monitors did not undergo regular re-calibration drifts in the observed data occurred imposing false trends which are small but in the same order of magnitude as the expected trends in tropospheric background ozone. This does not concern the ozone seasonality which we are particularly interested in.
The long-term data sets of agrometeorological variables including temperature and precipitation are available from September 1992 to the present day \citep[LandbruksMeteorologiske Tjeneste][note the station name here is Pasvik]{LMT_NIBIO}.

To relate local ozone concentration \chem{[O_3]} with visible damage on ozone sensitive plants, an ozone bio-monitor (referred to as ozone garden) has been installed at Svanhovd in the 2018/19 GS. The locations of the atmospheric monitoring site and the ozone garden are marked in the aerial photography shown in Fig.~\ref{fig:svanhovd_research_station}a. An ozone garden consists of selected plant species that are sensitive to ozone and are likely to display visible injuries. Cultivated species were, e.g. clovers (\emph{Trifolium repens}, \emph{Trifolium pratense}), tobacco (\emph{Nicotiana tabacum}, cultivars Bel-W3, Bel-B and Bel-C), and potato (\emph{Solanum tuberosum}). As shown in Fig.~\ref{fig:svanhovd_research_station}b, the plants had to be protected from herbivores with a wire-mesh fence. In 2018, we qualitatively observed visible ozone damage on semi-natural vegetation (clover) and crops (tobacco). In contrast, no such visible damage was found on the clovers in 2019, although it was observed on the sensitive tobacco cultivar (Bel-W3). Note that the extent of visible damage is not equivalent to a quantitative reduction of above/below ground biomass of a species.

@@

We determined the dominant species on-site at Svanhovd from Fig.~\ref{fig:svanhovd_research_station}a and found perennial grassland, birch (generalized as deciduous trees), and Scots pine (generalized as coniferous trees). Basic parameters for these PFTs are derived from \citet{EP:Simpson2007,GCB:Mills2011,ICP:MappingManual2017} and will be referred to as mapping manual (MM) parameterizations. The localization of parameters for coniferous trees is based on MM's Boreal Norway spruce, deciduous trees on MM's silver birch, and perennial grassland on MM's perennial grassland for Central Europe. For a comprehensive list of MM parameters, consult Supplement~Table~S1. In the following, we suggest localizations of key environmental parameters to a subarctic climate based on meteorological data to assess over-/underestimation of ozone damage when using MM Boreal parameterizations in ozone risk assessments.

@@

Along the conventional ozone monitoring at Svanhovd, we had installed a bio-monitoring with ozone sensitive species. Qualitatively, this showed visible ozone damage on clover and tobacco in 2018. In contrast, no such damage was found on the clovers in 2019. This might indicate a higher ozone damage risk under warmer climate conditions for these specific plant species in a subarctic climate. However, visible damage is not always correlated with a reduction in biomass \citep[and citations therein]{CRG:Felzer2007}. Also, our $\mathrm{DO_3SE}$ modeling results are in contrast to these observations at Svanhovd. Any visible damage caused by peak ozone flux through stomata or a specific dependency of plant sensibility to ozone on phenology stage are not accounted for in the $\mathrm{DO_3SE}$ model. In this regard, open-top chamber (OTC) experiments performed in northern Finland where Scots pine and downy birch seedlings were exposed to elevated ozone concentrations attributed a reduction in biomass and reproduction with visible damage explicitly to peak \chem{O_3} concentrations and fast phenological development at high growth rate \citep{Amb:Manninen2009}. Forbs and perennial grassland are more susceptible to ozone-induced damage in the reproductive state \citep{EP:Bassin2004}. 
That might explain why we observed damage in the ozone garden to a larger extent in 2018 than in 2019. An associated damage function to ozone depending on leaf age as proposed by \citet{AE:Musselman2006} or phenological state might improve the predictions.

>>>>

\subsubsection{Gap filling methodology}
\label{subsec:gap_filling}
The $\mathrm{DO_3SE}$ model requires hourly, continuous input data. Therefore, the following gap filling methodology was devised for the meteorological input data:
\begin{itemize}
\item Single hours of missing data were filled by taking the average of the hourly values before, and after, the missing value.
\item Several consecutive hours of missing data (23 or less) were filled by taking the average of the corresponding hour the day before, and the day after; and repeating this for each missing hour of data. If data were unavailable from that hour of the previous day, then only the value from the day after was used and vice versa. Interpolated values were not used in calculating averages.
\item Data gaps longer than $24\,\unit{h}$ were filled using weekly diurnal averages. i.e. an average was calculated using the corresponding hour throughout the week before and after. Interpolated values were not used in calculating averages.
\end{itemize}

@@

\subsubsection{Input data and observation site}
\label{subsec:data}
The required input data consists of ozone concentration $\chem{[O_3]}$, air temperature at $2\,\unit{m}$ $T_\mathrm{air}$, global irradiation $Q_0$, precipitation, $2\,\unit{m}$ wind $u_\mathrm{2m}$, and vapor pressure deficit VPD. VPD has been calculated from the Arden~Buck~equation~\citep{JAP:Buck1981, Buck2012}
\begin{equation}
  \mathrm{VPD} = P_s(T_\mathrm{air}) \cdot \left(1-\frac{\mathrm{relHum}}{100}\right),
\end{equation}
with saturation vapor pressure $P_s(T_\mathrm{air})$ in \unit{hPa}, observed $T_\mathrm{air}$, and relative humidity.

For our study focuses on ozone impact on subarctic vegetation, we chose Svanhovd in northern Norway ($69.45\,\unit{^\circ N}$, $30.03\,\unit{^\circ E}$, $30\,\unit{m\,a.s.l.}$) as site. Reasons for our choice were: (1) Ambient ozone concentration had been monitored there in the past, (2) an agrometeorological station is measuring, e.g. $2\,\unit{m}$ temperature, precipitation, global irradiance, and soil temperature at different depths, and (3) a national surveillance of air pollution monitoring (\chem{NO_x}, \chem{SO_x}, particular matter (PM), radiological species) is ongoing. Most of these measurements are performed by the Norwegian Institute for Air Research (NILU), who also conducted ozone measurements on our behalf during the 2018/19 growing season.

All essential climate variables for the 2018/19 growing season were provided by NILU through the web portal \href{luftkvalitet.no}{luftkvalitet.no} (last accessed May 2020). The \chem{O_3} monitoring data have been added to a database for air-quality monitoring called EBAS operated by NILU and are thus openly accessible. Long-term ozone observation data (1986--1996) were obtained from EBAS. Note that data from before the 1990s do not follow the high-quality standard procedures implemented nowadays and have to be treated with care \citep{NILU2003}. Because ozone monitors did not undergo regular re-calibration, drifts in the observed data occurred imposing false trends which are small but in the same order of magnitude as the expected trends in tropospheric background ozone. This does not concern the ozone seasonality which we are particularly interested in.
The long-term data sets of agrometeorological variables including temperature and precipitation are available from September 1992 to the present day \citep[LandbruksMeteorologiske Tjeneste][note the station name here is Pasvik]{LMT_NIBIO}.

To relate local ozone concentration \chem{[O_3]} with visible injuries on ozone sensitive plants, following \citet{GCB:Mills2011}, an ozone bio-monitor (referred to as ozone garden) has been installed at Svanhovd in the 2018/19 GS. The locations of the atmospheric monitoring site and the ozone garden are marked in the aerial photography shown in Fig.~\ref{fig:svanhovd_research_station}a. An ozone garden consists of selected plant species that are sensitive to ozone and are likely to display visible injuries. Cultivated species were, e.g. clovers (\emph{Trifolium repens}, \emph{Trifolium pratense}), tobacco (\emph{Nicotiana tabacum}, cultivars Bel-W3, Bel-B and Bel-C), and potato (\emph{Solanum tuberosum}). As shown in Fig.~\ref{fig:svanhovd_research_station}b, the plants had to be protected from herbivores with a wire-mesh fence. In 2018, we observed visible ozone injuries on semi-natural vegetation (clover) and crops (tobacco). In contrast, no such visible injuries were found on the clovers in 2019, although it was observed on the sensitive tobacco cultivar (Bel-W3). Note that the extent of visible injuries is, in general, not equivalent to a quantitative reduction of above/below ground biomass of a species.

@@

We determined the dominant species on-site at Svanhovd from Fig.~\ref{fig:svanhovd_research_station}a and found perennial grassland, birch (generalized as deciduous trees), and Scots pine (generalized as coniferous trees). Basic parameters for these PFTs are derived from \citet{EP:Simpson2007,GCB:Mills2011,ICP:MappingManual2017} and will be referred to as mapping manual (MM) parameterizations. The localization of parameters for coniferous trees is based on MM's Boreal Norway spruce, deciduous trees on MM's silver birch, and perennial grassland on MM's perennial grassland for Central Europe. For a comprehensive list of MM parameters, consult Supplement~Table~S1. In the following, we suggest localizations of key environmental parameters to a subarctic climate based on meteorological data to assess over-/underestimation of ozone-induced reduction of biomass when using MM Boreal parameterizations in ozone risk assessments.

@@

Along the conventional ozone monitoring at Svanhovd, we had installed a bio-monitoring with ozone sensitive species. Qualitatively, this showed visible ozone injuries on clover and tobacco in 2018. In contrast, no such damage was found on the clovers in 2019. This might indicate a higher ozone damage risk under warmer climate conditions for these specific plant species in a subarctic climate. However, visible injuries are not always correlated with a reduction in biomass \citep[and citations therein]{CRG:Felzer2007}. Also, our $\mathrm{DO_3SE}$ modeling results are in contrast to these observations at Svanhovd. Any visible injuries caused by peak ozone flux through stomata or a specific dependency of plant sensibility to ozone on phenology stage are not accounted for in the $\mathrm{DO_3SE}$ model. In this regard, open-top chamber (OTC) experiments performed in northern Finland where Scots pine and downy birch seedlings were exposed to elevated ozone concentrations attributed a reduction in biomass and reproduction with visible injuries explicitly to peak \chem{O_3} concentrations and fast phenological development at high growth rate \citep{Amb:Manninen2009}. Forbs and perennial grassland are more susceptible to ozone-induced damage in the reproductive state \citep{EP:Bassin2004}. 
That might explain why we observed visible injuries in the ozone garden to a larger extent in 2018 than in 2019. An associated damage function to ozone depending on leaf age as proposed by \citet{AE:Musselman2006} or phenological state might improve the predictions.
